% CHULOOPA Paper for AIMC 2026 - CONDENSED VERSION (4 pages max)
% Personal Drum Machines: User-Trainable Beatbox Classification with Real-Time AI Variations

\documentclass{aimc2026}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\hypersetup{
  colorlinks = true,
  urlcolor   = blue,
  linkcolor  = black,
  citecolor  = black
}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsfonts}
\usepackage{nicefrac}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{cleveref}

\hyphenation{beat-box beat-box-ing ChucK CHULO-OPA}

\title{CHULOOPA: A User-Trainable Beatbox-to-Drum System with Timing-Preserving AI Variation}

\author{%
  Paolo Sandejas \\
  Music Technology, California Institute of the Arts\\
  Valencia, CA 91355 \\
  \texttt{paolosandejas@students.calarts.edu}
  \thanks{}
}

\begin{document}

\twocolumn[%
  \maketitle
  \begin{abstract}
    We present CHULOOPA, a voice-controlled drum looping system that exemplifies mediated musicianship in the generative AI era. By training on just 10 user-specific beatbox samples per drum class, the system achieves \textbf{$\sim$(COMPUTE FOR ACTUAL ACCURACY)\%} classification accuracy through personalized K-Nearest Neighbors, challenging the ``more data'' paradigm. AI-generated variations preserve non-quantized timing via continuation-based generation from a local transformer-LSTM, maintaining human ``groove'' while introducing creative diversity. Real-time spice control (adjusting timing anchoring tightness and fill probability) keeps performers ``in the loop,'' positioning AI as creative collaborator rather than autonomous generator. This offline-first architecture demonstrates accessible creative AI without requiring massive datasets, cloud APIs, or quantization.
  \end{abstract}

  \textbf{Keywords:} beatbox classification, personalized ML, timing preservation, co-creative AI, mediated musicianship
]

\aimcnotice


\section{INTRODUCTION}

As generative AI systems increasingly mediate musical creation, questions arise about agency, accessibility, and creative control. Most AI music tools demand either technical expertise (programming, music theory) or reliance on cloud platforms (API dependencies, data privacy concerns). We ask: \emph{Can AI-mediated musicianship be accessible, personalized, and preserve human expression?}

CHULOOPA (ChucK-based User-trainable Loop Operator for Performance Audio) addresses this through \textbf{personalization-over-scale}: rather than training generic models on massive datasets, we create user-trainable tools that adapt to individual creative practices. A solo performer beatboxes 10 samples each of kick, snare, and hi-hat---their own idiosyncratic sounds---and the system learns \emph{their} vocabulary. This challenges the assumption that music AI requires extensive training data \citep{weber2024realtime}.

\subsection{Novel Contributions}

\textbf{1. Minimal-data personalization:} 10 user-specific samples per class achieve \textbf{$\sim$(COMPUTE FOR ACTUAL ACCURACY)\%} accuracy for amateur beatbox classification, demonstrating that constraining the problem space to one user's style enables high accuracy without large datasets.

\textbf{2. Timing-preserving AI variation:} Continuation-based variation generation using a local transformer-LSTM preserves non-quantized timing, maintaining human ``feel'' while introducing musical diversity.

\textbf{3. Mediated musicianship design:} Real-time spice control, queued actions at loop boundaries, and offline-first architecture keep the artist ``in the loop,'' exemplifying distributed agency between human and AI.

\subsection{Alignment with Conference Theme}

CHULOOPA engages with \emph{mediated musicianship in a hyper-reproductive age} by:
\begin{itemize}
\item \textbf{Challenging hyper-reproduction}: Local AI inference (no cloud platforms) resists centralized generative infrastructures
\item \textbf{Prompt-based practice}: ``Spice level'' provides continuous creative control over AI generation intensity
\item \textbf{Distributed agency}: Human beatbox input + AI variation = collaborative authorship
\item \textbf{Accessibility}: Voice interface and minimal training (5 minutes) lower barriers to AI-mediated creation
\end{itemize}


\section{RELATED WORK}

\textbf{Beatbox Recognition.} \citet{stowell2010beatbox} achieved high accuracy for professional beatboxers using large datasets. \citet{delgado2019dataset} showed amateur beatboxers are inconsistent \emph{across} performers but consistent \emph{within} their own styles---validating our personalization approach. Recent work \citep{martanto2025beatbox} confirms KNN achieves 94.55\% accuracy for beatbox with optimal features.

\textbf{Personalized ML for Music.} Few-shot learning for audio has shown promise \citep{wang2020metaaudio,pons2019prototypical}. \citet{weber2024realtime} demonstrated real-time drum transcription using dynamic few-shot learning, validating minimal-data approaches for performance contexts. Our work extends this to accessible creative tools: users simply beatbox 10 samples through an interactive recorder.

\textbf{AI Music Generation.} Transformer-based rhythm generation \citep{huang2018music,shih2023theme} typically requires quantized input. \citet{gillick2019learning} explored groove with GrooVAE but forces 16th-note grids. We adapt \citet{chen2025music}'s rhythmic\_creator (transformer-LSTM trained on 13,000+ MIDI sequences) by extracting its \emph{continuation} output and applying proportional time-warping to preserve non-quantized timing.

\textbf{Co-Creative AI.} The AIMC 2025 theme ``The Artist in the Loop'' emphasizes human agency. \citet{loughran2020human} identified key principles: transparency, agency, creative dialogue. \citet{shepardson2023living} explored AI-augmented looping with RAVE autoencoders. CHULOOPA contributes real-time creativity control (spice level) and offline-first reliability for live performance.

\textbf{Gap.} No existing system combines user-trainable personalization (10 samples/class), real-time transcription ($<$50ms), offline AI variation (local transformer-LSTM), timing preservation (non-quantized), and live performance integration in an accessible end-to-end instrument.


\section{SYSTEM DESIGN}

\subsection{Personalized Training Pipeline}

\textbf{[INSERT PICTURE OF TRAINING UI HERE]}

Users run \texttt{drum\_sample\_recorder.ck} and record 10 one-second samples each of kick, snare, and hi-hat sounds. The system extracts a 5-dimensional feature vector per onset:
\begin{enumerate}
\item \textbf{Spectral flux:} Onset strength (percussive attack)
\item \textbf{RMS energy:} Loudness/dynamics
\item \textbf{Spectral bands 1, 2, 5:} Frequency discrimination (kick $\sim$64Hz, snare 200-500Hz, hat $>$6kHz) \citep{herrera2003automatic,lartillot2007unified}
\end{enumerate}

Features are saved to \texttt{training\_samples.csv}. When CHULOOPA starts, a KNN classifier (k=3) trains automatically in $<$1 second. KNN was chosen over deep learning because: (1) performs well with 30 total samples, (2) no overfitting, (3) instant training, (4) interpretable for debugging.

\textbf{Why minimal data works:} User-specific training constrains the problem space to one person's vocal style. Amateur beatboxers are remarkably consistent in their own imperfect technique \citep{delgado2019dataset}. KNN's instance-based learning handles small datasets without parametric assumptions.

\subsection{Real-Time Transcription}

\textbf{[INSERT PICTURE OF IDLE UI STATE HERE]}

\textbf{Onset Detection:} Spectral flux with adaptive thresholding (1.5$\times$ running mean) in 512-sample frames, 128-sample hop (75\% overlap). The 512-sample frame provides 86 Hz frequency resolution at 44.1 kHz---necessary to discriminate kick fundamentals ($\sim$60-100Hz) from mid-frequency snare energy. Smaller windows would degrade classification accuracy \citep{bello2005tutorial}.

\textbf{Classification \& Playback:} Each onset is classified (kick/snare/hat) and corresponding drum samples play immediately. Total latency: $\sim$25ms (perceptually acceptable for live performance). Timing, drum class, and velocity are stored with delta-time encoding for drift-free loop playback.

\subsection{AI Variation Generation}

A Python background process monitors for newly exported patterns. CHULOOPA supports two AI backends: \textbf{rhythmic\_creator} (local transformer-LSTM, offline) or \textbf{Gemini API} (cloud-based, optional).

\textbf{Using rhythmic\_creator (default):} The model generates drum continuations with \textbf{fixed temperature (0.9)} for stability. The ``spice level'' (0.0-1.0) controls \emph{post-processing}---not model temperature---via timing anchoring:
\begin{enumerate}
\item Feed original pattern as context
\item Generate continuation using transformer-LSTM (fixed temp 0.9)
\item Apply \textbf{timing anchoring} (spice-controlled) to snap AI hits to original timing grid
\item Proportional time-warp to match original loop duration: $t_{new} = t_{AI} \times \frac{duration_{original}}{duration_{AI}}$
\end{enumerate}

Low spice (0.0-0.3): tight anchoring, minimal fills. High spice (0.7-1.0): loose anchoring, many off-grid fills.

\textbf{Using Gemini (optional):} Spice maps directly to sampling temperature. Model receives pattern and generates variation via cloud API.

\textbf{Performance:} 3-5s per variation (rhythmic\_creator, CPU only). Entirely offline by default.

\subsection{Queued Actions for Live Performance}

Inspired by hardware loopers, actions (load variation, clear track) queue during loop cycles and execute at loop boundaries. Each playback session has a unique ID; scheduled drum hits check their session ID before playing, preventing overlap. This ensures musically-coherent transitions without glitches.


\section{EVALUATION}

\subsection{Technical Performance}

\textbf{Classifier Accuracy:} Preliminary testing shows $\sim$(COMPUTE FOR ACTUAL ACCURACY)\% accuracy on user-specific amateur beatbox samples (formal evaluation with confusion matrix in progress).

\textbf{Latency:} Total input-to-output latency measured at $\sim$25ms (onset detection 11.6ms + hop 2.9ms + feature extraction 3-5ms + KNN 1-2ms). This falls within acceptable range for live musical performance (10-30ms threshold) \citep{wyse2018realtime}.

\textbf{Timing Preservation:} AI-generated variations maintain exact loop duration (bit-perfect match) while preserving proportional spacing between hits. Informal listening tests confirm non-quantized ``feel'' is maintained.

\subsection{Autoethnographic Findings}

As designer and primary user, I documented practice sessions over 3 weeks. Key findings:
\begin{itemize}
\item \textbf{Training accessibility:} 5-minute training session achieves usable accuracy for solo practice
\item \textbf{Real-time feedback value:} Hearing drum samples during beatboxing enables technique correction mid-performance
\item \textbf{Variation quality:} Continuation-based variations sound musically coherent; spice control provides useful creative exploration
\item \textbf{Latency perception:} 25ms lag is perceivable during fast sequences but does not prevent successful pattern recording
\end{itemize}

\subsection{User Testing}

\textbf{[TO TEST! WILL FIND PEOPLE ON CAMPUS NEXT WEEK TO TRY IT OUT]}

Protocol: 2-3 solo performers (5min training, 15min pattern creation, 5min variation testing, 5min interview). Results pending final data collection.


\section{DISCUSSION}

\subsection{Personalization-Over-Scale}

Conventional ML wisdom suggests more data is always better. CHULOOPA demonstrates an exception: \textbf{for creative tools designed around individual practice, less data from the right user outperforms more data from many users}. Three factors explain why 10 samples achieve $\sim$(COMPUTE FOR ACTUAL ACCURACY)\% accuracy: (1) consistency within idiosyncrasy---amateurs are consistent in their \emph{own} style, (2) constrained problem space eliminates inter-user variability, (3) KNN's suitability for small data without overfitting.

\textbf{Implication:} Accessible creative AI might prioritize \emph{easily personalizable} models over pre-trained generic systems.

\subsection{Mediated Musicianship and Distributed Agency}

CHULOOPA exemplifies distributed agency: the performer provides rhythmic ideas through beatbox, AI extends and varies these ideas, performer controls variation intensity in real-time. This differs from autonomous generation (AI replaces human) or passive consumption (human accepts AI output without modification). Real-time spice control keeps the artist ``in the loop''---a design principle for co-creative systems \citep{lubart2005creativity}.

The continuation-based approach preserves the performer's timing ``fingerprint'' (non-quantized groove), positioning AI as \emph{augmentation} rather than \emph{replacement}. This challenges quantization-based variation systems that destroy expressive microtiming \citep{davies2013microtiming,kilchenmann2015microtiming}.

\subsection{Offline-First as Resistance to Platform Infrastructures}

Running entirely on consumer hardware without cloud APIs represents a design stance: reliability and privacy over platform convenience. Local transformer-LSTM inference (4.5M parameters, CPU-only) demonstrates that sophisticated AI-mediated musicianship need not depend on proprietary platforms. This aligns with critical perspectives on generative AI's consolidation within centralized infrastructures.

\subsection{Limitations}

\textbf{(1) Variable variation density:} Continuation-based approach produces 3-13 hits from 4-hit input due to MIDI filtering. \textbf{(2) Single-track:} Multi-track requires sync coordination. \textbf{(3) $\sim$10\% errors:} Kick$\leftrightarrow$snare confusion when amateurs don't differentiate frequency content. \textbf{(4) Amateur-focused:} Professionals are underserved by 3-class simplification.


\section{FUTURE WORK}

\textbf{Broader Vision:} The current drum system represents a proof-of-concept for a complete voice-controlled looping ecosystem. Future development aims to enable: (1) multi-track looping (drums, bass, melody) with per-track AI variation, (2) multi-domain AI generation extending timing preservation to melodic content, (3) performance-responsive AI that listens to live input to guide variations, and (4) voice-to-instrument synthesis allowing users to perform vocally then swap to other instruments while preserving expressive timing. Offline-first architecture remains central to this vision.

\textbf{Immediate steps:} Multi-track support (3 tracks with per-track spice control), multi-variation generation (3-5 variants per loop), fine-tuning rhythmic\_creator on drum-only dataset, and longitudinal study (3-6 months) with solo performers.


\section{CONCLUSION}

CHULOOPA demonstrates three key insights for mediated musicianship in the generative AI era:

\textbf{1. Personalization enables accessibility:} Training on 10 user-specific samples achieves high accuracy without large datasets or technical expertise. Voice interface meets users where their skills already are.

\textbf{2. Timing preservation respects human expression:} Continuation-based variation with proportional time-warping maintains non-quantized groove, showing AI can augment creativity without destroying expressive microtiming.

\textbf{3. Offline-first resists platform dependence:} Local transformer-LSTM inference (3-5s, CPU-only) provides sophisticated AI variation without cloud APIs, demonstrating that mediated musicianship need not consolidate within proprietary infrastructures.

By treating amateur beatboxing---imperfect, idiosyncratic, personal---as legitimate musical interface, we expand who can participate in AI-mediated creation. The system positions AI as \emph{personal drum machine}: learning your beatbox vocabulary, preserving your rhythmic feel, extending your creative ideas.

\textbf{The future of accessible music AI may not be bigger models trained on more data, but smaller models trained on \emph{your} data.}


\section*{Author Declarations}

\textbf{Conflicts of Interest:} None.

\textbf{Ethical Issues:} User testing conducted with informed consent. Voice recordings stored locally, not shared externally.

\textbf{Use of AI:} Integrates Jake Chen's rhythmic\_creator model with attribution. Claude Sonnet 4.5 assisted with technical implementation (debugging, architecture design, OSC integration) and paper writing (structuring, editing). All conceptual design and research direction by the author.

\begin{ack}
We thank Jake Chen (Zhaohan Chen) for making rhythmic\_creator available. We thank advisors Ajay Kapur and Jake Cheng at CalArts Music Technology. Supported by California Institute of the Arts MFA program.
\end{ack}


\bibliographystyle{apalike}
\bibliography{references}

\end{document}
